Spike-and-Slab Generalized Additive Models and Scalable Algorithms for High-Dimensional Data

### 2022-02-21 Manuscript/03-Simulation.Rmd
Add prediction result first draft

To evaluate the predictive performance of the models, the statistics, $R^2$ for Gaussian model and AUC for binomial model calculated based on the testing dataset, are averaged <!--and deviance are calculated for Poisson model,--> across 50 simulations. \bg{To evaluate the variable selection performance of the models, we record the set of variables each model selects and calculate the averaged false discover rate (FDR), false negative rate(FNR), and Matthews correlation coefficient (MCC). For method that doesn't automatically achieve variable selection, we set alpha level at 0.05 for mgcv that relies on hypothesis testing for variable selection, and soft-threshold at 0.5 for spikeSlabGAM given the marginal inclusion probabilities. For the two methods, BHAM and spikeSlabGAM, that are capable of bi-level selection, we record the probability that the linear and nonlinear components of each predictors are selected in the models.
MCC = \frac{TP\times TN - FP \times FN}{\sqrt{(TP+FP)(TP+FN)(TN+FP)(TN+FN)}},
$$
where TP, TN, FP, and FN are true positives, true negatives, false positives, and false negatives, respectively.
$$
}
\bg{\subsection{Simulation Results}}
\bg{\subsubsection{Prediction Performance}}
The predictive performances have a consistent pattern across the two distributions of outcomes. Across all the scenarios, COSSO and adaptive COSSO have the least favorable performance among the applicable methods examined (See Table \ref{tab:gaus} and \ref{tab:bin_auc}). To note, mgcv doesn't support high-dimensional analysis, i.e. the number of coefficients are greater than the sample size, and hence not evaluated when $p=100, 200$. mgcv predicts well when $p$ is small or moderate ($p = 4, 10$), and deteriorate when the number of predictors increase. Among the three fast-computing Bayesian hierarchical models, the proposed models, BHAM-IWLS and BHAM-CD predicts better than SB-GAM when the dimension of are moderate (p=4, 10, 50). Particularly, BHAM-IWLS performs as good as mgcv if not better. However, in high dimensional case where we mimic the situation the signals are extremely sparse, SB-GAM has better performance than the proposed method. However, the BHAM-CD has extreme computational advantage over SB-GAM (see Table \ref{tab:time_sim}) without sacrificing much of the prediction accuracy.

\bg{

\bg{\subsubsection{Variable Selection Performance}
\begin{itemize}
\item If we use mcc as the measure, SBGAM has the best performance among all the method of comparison, followed by the proposed model BHAM. The pattern is consitent regardless of the type of outcomes. To note here, mgcv is not included in the performance table, as the mgcv isn't compatable with hign-dimensitonal data. Nevertheless, when the dimensionality is low, mgcv perfoms while for variable selection.
\item If we look at the precision of the variable selection, BHAM performs better than SBGAM model particularaly when the dimensionality increases. While the recall is lower than SBGAM. This translate to the selection of BHAM is more conservative that SBGAM, such that among the selected variables, many of them are signals, however, due to the conversative selection, there are few variables are selected.
\item We also want to high light that in the set of linear simulation, when the outcome is cgaussina, the proposed model performed extremely while. On the contrary the SBGAM is not so well.
\end{itemize}

>>>>

To evaluate the predictive performance of the models, the statistics, $R^2$ for Gaussian model and AUC for binomial model calculated based on the testing dataset, are averaged <!--and deviance are calculated for Poisson model,--> across 50 simulations. \bg{To evaluate the variable selection performance of the models, we record the set of variables each model selects and calculate the averaged false discover rate (FDR), false negative rate(FNR), and Matthews correlation coefficient (MCC). For method that doesn't automatically achieve variable selection, we set alpha level at 0.05 for mgcv that relies on hypothesis testing for variable selection, and soft-threshold at 0.5 for spikeSlabGAM given the marginal inclusion probabilities. For the two methods, BHAM and spikeSlabGAM, that are capable of bi-level selection, we record the probability that the linear and nonlinear components of each predictors are selected in the models.
$$
MCC = \frac{TP\times TN - FP \times FN}{\sqrt{(TP+FP)(TP+FN)(TN+FP)(TN+FN)}},
$$
where TP, TN, FP, and FN are true positives, true negatives, false positives, and false negatives, respectively.

}
\bg{\subsection{Simulation Results}}
\bg{\subsubsection{Prediction Performance}}
The predictive performances have a consistent pattern across the two distributions of outcomes. Across all the scenarios, COSSO and adaptive COSSO have the least favorable performance among the applicable methods examined (See Table \ref{tab:gaus} and \ref{tab:bin_auc}). To note, mgcv doesn't support high-dimensional analysis, i.e. the number of coefficients are greater than the sample size, and hence not evaluated when $p=100, 200$. mgcv predicts well when $p$ is small or moderate ($p = 4, 10$), and deteriorate when the number of predictors increase. Among the three fast-computing Bayesian hierarchical models, the proposed models, BHAM-IWLS and BHAM-CD predicts better than SB-GAM when the dimension of are moderate (p=4, 10, 50). Particularly, BHAM-IWLS performs as good as mgcv if not better. However, in high dimensional case where we mimic the situation the signals are extremely sparse, SB-GAM has better performance than the proposed method. However, the BHAM-CD has extreme computational advantage over SB-GAM (see Table \ref{tab:time_sim}) without sacrificing much of the prediction accuracy.

\bg{
\begin{itemize}
  \item When the linear assumptions are violated, the linear lasso model performs extremely bad. adaptive cosso have improvement upon cosso.
  \item To note, the drastic deteriation in the prediction of mgcv is caused by the model would not converge in those scenarios. A null model were returned as a final model. 
  \item Within the simulations where the functional form of the predictors are linear, we see that the proposed model similar to the lasso model regardless of the type outcome, and better moderately better than the model of comparison. (See Table Lnr Gause, and supporting table Lnr_binom.) Particularly, in gaussina simulations, the proposed model shows great advantage compared to SB_GAM (numeralize the advantage), and slightly improvement compare to ssGAM (numeralize the advantage). This shows that the model can be applied without constraining to the assumption where all functional form are non-linear.
  \item

\end{itemize}
}

\bg{\subsubsection{Variable Selection Performance}
While the main utility of the proposed work is prediction, the proposed model can also serve as variable selection/functional selection as a by-product. Hence We evaluate the variable selection performance.

\begin{itemize}
\item If we use mcc as the measure, SBGAM has the best performance among all the method of comparison, followed by the proposed model BHAM. The pattern is consitent regardless of the type of outcomes. To note here, mgcv is not included in the performance table, as the mgcv isn't compatable with hign-dimensitonal data. Nevertheless, when the dimensionality is low, mgcv perfoms while for variable selection.
\item If we look at the precision of the variable selection, BHAM performs better than SBGAM model particularaly when the dimensionality increases. While the recall is lower than SBGAM. This translate to the selection of BHAM is more conservative that SBGAM, such that among the selected variables, many of them are signals, however, due to the conversative selection, there are few variables are selected.
\item We also want to high light that in the set of linear simulation, when the outcome is cgaussina, the proposed model performed extremely while. On the contrary the SBGAM is not so well.
\end{itemize}

### 2022-02-21 Manuscript/SS_GAM.tex
Add prediction result first draft

In this section, we compare the performance of the proposed models to
four alternative models: component selection and smoothing operator
(COSSO) \citep{Zhang2006GAM}, adaptive COSSO \citep{Storlie2011},
generalized additive models with automatic smoothing \citep{Wood2011},
SB-GAM
\citep{Bai2021}\textcolor{red}{, linear lasso model and spikeSlabGAM [TODO: add citation here]}.
COSSO is one of the earliest smoothing spline models that consider
sparsity-smoothness penalty. Adaptive COSSO improved upon COSSO by using
adaptive weight for penalties such that the penalty of each functional
component are different for extra flexibility. Generalized additive
models with automatic smoothing, hereafter \textit{mgcv}, is one of the
most popular models for nonlinear effect interpolation and prediction.
SB-GAM is the first spike-and-slab lasso GAM. We implement COSSO and
adaptive COSSO with R package \texttt{cosso 2.1-1}, generalized additive
models with automatic smoothing with R package \texttt{mgcv 1.8-31},
SB-GAM with R package \texttt{sparseGAM 1.0}. COSSO models and SB-GAM do
not provide flexibility to define smoothing functions, and hence use the
default choices. Both mgcv and proposed models allow customized
smoothing functions and we choose the cubic regression spline. We
controll the dimensionality of each smoothing function, 10 bases, for
all different choices of smoothing functions. We use 5-fold CV with the
default selection criteria to select the final model for COSSO models,
SB-GAM and the proposed models. 20 default candidates of tuning
parameters (\(s_0\) in BHAM, \(\lambda_0\) in SB-GAM) are examined for
SB-GAM and the proposed models which allow user-specification of tuning
candidates. All computation was conducted on a high-performance 64-bit
Linux platform with 48 cores of 2.70GHz eight-core Intel Xeon E5-2680
processors and 24G of RAM per core and R 3.6.2 \citep{R}.

@@

To evaluate the predictive performance of the models, the statistics,
\(R^2\) for Gaussian model and AUC for binomial model calculated based
on the testing dataset, are averaged across 50 simulations.
\textcolor{red}{To evaluate the variable selection performance of the models, we record the set of variables each model selects and calculate the averaged false discover rate (FDR), false negative rate(FNR), and Matthews correlation coefficient (MCC). For method that doesn't automatically achieve variable selection, we set alpha level at 0.05 for mgcv that relies on hypothesis testing for variable selection, and soft-threshold at 0.5 for spikeSlabGAM given the marginal inclusion probabilities. For the two methods, BHAM and spikeSlabGAM, that are capable of bi-level selection, we record the probability that the linear and nonlinear components of each predictors are selected in the models.}

\begin{center}
Table \ref{tab:gaus} here\\
Table \ref{tab:bin_auc} here\\
\end{center}

\textcolor{red}{\subsection{Simulation Results}}
\textcolor{red}{\subsubsection{Prediction Performance}}

The predictive performances have a consistent pattern across the two
distributions of outcomes. Across all the scenarios, COSSO and adaptive
COSSO have the least favorable performance among the applicable methods
examined (See Table \ref{tab:gaus} and \ref{tab:bin_auc}). To note, mgcv
doesn't support high-dimensional analysis, i.e.~the number of
coefficients are greater than the sample size, and hence not evaluated
when \(p=100, 200\). mgcv predicts well when \(p\) is small or moderate
(\(p = 4, 10\)), and deteriorate when the number of predictors increase.
Among the three fast-computing Bayesian hierarchical models, the
proposed models, BHAM-IWLS and BHAM-CD predicts better than SB-GAM when
the dimension of are moderate (p=4, 10, 50). Particularly, BHAM-IWLS
performs as good as mgcv if not better. However, in high dimensional
case where we mimic the situation the signals are extremely sparse,
SB-GAM has better performance than the proposed method. However, the
BHAM-CD has extreme computational advantage over SB-GAM (see Table
\ref{tab:time_sim}) without sacrificing much of the prediction accuracy.

\textcolor{red}{\subsubsection{Variable Selection Performance}}

>>>>

In this section, we compare the performance of the proposed models to
four alternative models: component selection and smoothing operator
(COSSO) \citep{Zhang2006GAM}, adaptive COSSO \citep{Storlie2011},
generalized additive models with automatic smoothing \citep{Wood2011},
SB-GAM
\citep{Bai2021}\textcolor{red}{, linear lasso model and spikeSlabGAM [TODO: add citation here].[TODO: explain about linear lasso model and spikeSlabGAM model]}.
COSSO is one of the earliest smoothing spline models that consider
sparsity-smoothness penalty. Adaptive COSSO improved upon COSSO by using
adaptive weight for penalties such that the penalty of each functional
component are different for extra flexibility. Generalized additive
models with automatic smoothing, hereafter \textit{mgcv}, is one of the
most popular models for nonlinear effect interpolation and prediction.
SB-GAM is the first spike-and-slab lasso GAM. We implement COSSO and
adaptive COSSO with R package \texttt{cosso 2.1-1}, generalized additive
models with automatic smoothing with R package \texttt{mgcv 1.8-31},
SB-GAM with R package \texttt{sparseGAM 1.0}. COSSO models and SB-GAM do
not provide flexibility to define smoothing functions, and hence use the
default choices. Both mgcv and proposed models allow customized
smoothing functions and we choose the cubic regression spline. We
controll the dimensionality of each smoothing function, 10 bases, for
all different choices of smoothing functions. We use 5-fold CV with the
default selection criteria to select the final model for COSSO models,
SB-GAM and the proposed models. 20 default candidates of tuning
parameters (\(s_0\) in BHAM, \(\lambda_0\) in SB-GAM) are examined for
SB-GAM and the proposed models which allow user-specification of tuning
candidates. All computation was conducted on a high-performance 64-bit
Linux platform with 48 cores of 2.70GHz eight-core Intel Xeon E5-2680
processors and 24G of RAM per core and R 3.6.2 \citep{R}.

@@

To evaluate the predictive performance of the models, the statistics,
\(R^2\) for Gaussian model and AUC for binomial model calculated based
on the testing dataset, are averaged across 50 simulations.
\textcolor{red}{To evaluate the variable selection performance of the models, we record the set of variables each model selects and calculate the averaged false discover rate (FDR), false negative rate(FNR), and Matthews correlation coefficient (MCC). For method that doesn't automatically achieve variable selection, we set alpha level at 0.05 for mgcv that relies on hypothesis testing for variable selection, and soft-threshold at 0.5 for spikeSlabGAM given the marginal inclusion probabilities. For the two methods, BHAM and spikeSlabGAM, that are capable of bi-level selection, we record the probability that the linear and nonlinear components of each predictors are selected in the models.
MCC = \frac{TP\times TN - FP \times FN}{\sqrt{(TP+FP)(TP+FN)(TN+FP)(TN+FN)}},
$$
where TP, TN, FP, and FN are true positives, true negatives, false positives, and false negatives, respectively.
$$
} The predictive performances have a consistent pattern across the two
distributions of outcomes. Across all the scenarios, COSSO and adaptive
COSSO have the least favorable performance among the applicable methods
examined (See Table \ref{tab:gaus} and \ref{tab:bin_auc}). To note, mgcv
doesn't support high-dimensional analysis, i.e.~the number of
coefficients are greater than the sample size, and hence not evaluated
when \(p=100, 200\). mgcv predicts well when \(p\) is small or moderate
(\(p = 4, 10\)), and deteriorate when the number of predictors increase.
Among the three fast-computing Bayesian hierarchical models, the
proposed models, BHAM-IWLS and BHAM-CD predicts better than SB-GAM when
the dimension of are moderate (p=4, 10, 50). Particularly, BHAM-IWLS
performs as good as mgcv if not better. However, in high dimensional
case where we mimic the situation the signals are extremely sparse,
SB-GAM has better performance than the proposed method. However, the
BHAM-CD has extreme computational advantage over SB-GAM (see Table
\ref{tab:time_sim}) without sacrificing much of the prediction accuracy.

\textcolor{red}{

}

\textcolor{red}{\subsubsection{Variable Selection Performance}
\begin{itemize}
\item If we use mcc as the measure, SBGAM has the best performance among all the method of comparison, followed by the proposed model BHAM. The pattern is consitent regardless of the type of outcomes. To note here, mgcv is not included in the performance table, as the mgcv isn't compatable with hign-dimensitonal data. Nevertheless, when the dimensionality is low, mgcv perfoms while for variable selection.
\item If we look at the precision of the variable selection, BHAM performs better than SBGAM model particularaly when the dimensionality increases. While the recall is lower than SBGAM. This translate to the selection of BHAM is more conservative that SBGAM, such that among the selected variables, many of them are signals, however, due to the conversative selection, there are few variables are selected.
\item We also want to high light that in the set of linear simulation, when the outcome is cgaussina, the proposed model performed extremely while. On the contrary the SBGAM is not so well.
\end{itemize}

}
